paper2_resistance_mechanisms_mixtures

### 2016-09-14 inst/documents/paper2_resistance_mechanisms_mixtures.Rmd
paper writing

### version 3

# Mechanisms by which insecticide resistance can be influenced by strategies of insectcide use : a simple theoretical exploration.

@@

## ANDY TO DO
1. work out what's going on with exposure ... why does increasing exposure have a greater effect on mixture than sequential ? Is it because the exposure to one insecticide also effects the development of resistance in the other ?
1. check & add to discussion text
1. add to intro about results in levick et al. 
1. explain resistance-restoration and its relation to selection coefficient

@@

This paper explores the mechanisms by which insectide resistance is expected to increase with the use of insecticides in isolation, sequence/rotation and mixtures. It is aimed at a readership with limited experience of population genetics and resistance. It uses a model representing two independent loci coding for resistance to two insecticides. It explores factors that are expected to influence the development of resitance, notably the starting frequency of resistance alleles, their dominance, the exposure of insects to the insecticides, the effectiveness of the insecticides, the selective advantage of resistance and the dominance of the alleles. There is uncertainty about the values of these parameters in field situations. We demonstrate how the values of these parameters are expected to influence the development of resistance and the relative benefits of alternative insecticide application strategies. We describe an R package and user-interface that allows further exploration of the development of resistance. The user-interface (https://andysouth.shinyapps.io/MixSeqResist1) allows replication of our results and exploration of what happens when inputs are changed outside of the ranges we have used.

@@

@@

We use a 2-locus population genetic model of insecticide resistance (Levick et al. in press) to illustrate the mechanisms by which these parameters are expected to influence the development of insectide resistance under alternative insecticide use strategies. Levick et al. (in press) describe the technical details of the model and explore large parameter ranges. In contrast we provide an accessible summary of the model and use selected parameter values to demonstrate mechanisms that can influence the development of resistance.

@@

*start with the framework, then after describe specifics about how many loci etc*
*try hard to start simple, explain all terms and make few assumptions*

@@

The simulation proceeds through generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness. This acts to make the fitter alleles more common over time. Sexual reproduction is represented first by creating gametes by splitting the adult genomes in two, and secound by forming offspring genotypes by adding gametes with recombination.

@@

Here we principally look at model outputs of the change in allele frequencies over time (generations). We investigate three insecticide use strategies.
1. single insecticide
1. two insecticides used in sequence, replacing the first with a second once the proportion of resistance alleles reaches 50%
1. two insecticides used in a mixture (concurrently)

@@

*? it would be nice to add grey lines to fig6 & others for the 'other' scenario to facilitate comparison, but I'd need to do some tricky rewriting of Beths code.*

@@

#### results: Exposure
Increasing exposure to 0.8 reduces time-to-resistance for both mixture and sequential scenarios (Fig 8A relative to 4A). The change in exposure has a greater effect on the mixture relative to the sequential strategy and this results in a greater relative benefit for sequence versus mixture. Remembering that earlier increasing one effectiveness to 0.8 resulted in mixtures being favoured over sequence (Fig 6B). Increasing exposure from 0.5 to 0.8 switches back to sequential being the favoured strategy (Fig 8B). This demonstrates how higher exposures can favour sequential use.

@@

When single insecticide use was represented (Figs 3-5) in all cases the initial increase in resistance was most rapid with straight lines on the log plots up to around 50% followed by a slowing rate of increase up to 100%. Higher values of the five inputs invetsigated all led to more rapid increases in resistance.

The response of resistance to changing exposure and effectiveness was identical (compare Fig 3A to 3B). This makes sense as vector kill is effectively a product of exposure times effectiveness. For example exposing 50% of a  population to an insecticide which is 75% effective would be expecetd to have the same result as exposing 75% of the population to one which is 50% effective. This observation points to the mechanism by which increasing both exposure and effectiveness lead to a faster increase in resistance. In both cases the increased deaths of susceptible vectors cause a higher selection pressure that can explain the faster increase.

*can we explain why at low dominance 0.2 the rate of increase in resistance is slower at the start ?*

@@

In comparison with the use of a single insecticide, using two insecticides either in sequence or in mixtures differs in some ways and where it differs that can again help us to understand the mechanisms. Looking at our base scenario (Fig 6A)

@@

*may want to copy or move some text from summary points below into discussion here*

@@

(@) In a mixture the effectiveness of one insecticide has major effects on the resistance of the other. In contrast the dominance and selective advantage of one insecticide has less effect on the resistance of the other. For exposure we have kept values the same for each insecticide the same (to represent a mixture in which the 2 insecticides are applied together). However we predict that if they were applied separately then (similar to effectiveness) the level of exposure to one insecticide would influence the time-to-resistance of the other.

@@

(@) Increasing exposure gives shorter time-to-resistance for both mixtures and sequences due to increased deaths of susceptibles creating a higher selection pressure. In a mixture the effect of exposure is greater and this means that increasing exposure can make sequences preferable to mixtures. The reason for the greater effect of exposure on mixtures could be that exposure is having both a direct effect on each insecticide and an indirect effect on the other. Thus higher exposures lead to faster resistance in each insecticide (direct), in turn increased resistance to each insecticide reduces the 'protection' that it gives the other, leading to faster resistance in the other (indirect). This could act as a postive feedback loop where faster resistance in one contributes to faster resistance in the other.

@@

1. In discussion and maybe elewhere address potential concern that the relationships shown here may not hold for all of paramter space.
1. can I develop explanatory diagrams to describe the input parameters ?
1. mention generation time of 1 month
1. mention other potential effects e.g. differential exposure of M&F, sex linkage, ...
1. add to discussion about relevance of these mechanisms in the field
1. e.g. exposure can be reduced if either insecticide has a repellant effect. (Birget & Koella 2015 mention this not for resistance but for ITNs with repellancy requiring higher coverage for elimination)

>>>>

### version 5

# Mechanisms by which insecticide resistance can be influenced by strategies of insecticide use : exploring by simulation.

@@

## ANDY TO DO
1. work out what's going on with exposure ... why does increasing exposure have a greater effect on mixture than sequential ? 
1. add to intro including about about results in levick et al. 
1. explain resistance-restoration and its relation to selection coefficient
1. expand discussion to include relevance to real world

@@

This paper explores the mechanisms by which insectide resistance is expected to increase with the use of insecticides in isolation, sequence and mixtures. It is aimed at a readership with limited experience of population genetics and resistance. It uses a model representing two independent loci coding for resistance to two insecticides. It explores factors that are expected to influence the development of resitance, notably the starting frequency of resistance alleles, their dominance, the exposure of insects to the insecticides, the effectiveness of the insecticides, and the selective advantage of resistance. We use the model to develop a mechanistic understanding of how the values of these parameters are expected to influence the development of resistance and the relative benefits of alternative insecticide application strategies. We expect this mechanistic understanding to be more robust to uncertainty in parameter values in the field. We describe an R package and user-interface that allows further exploration of the development of resistance. The user-interface (https://andysouth.shinyapps.io/MixSeqResist1) allows replication of our results and exploration of what happens when inputs are changed outside of the ranges we have used.

@@

We use a 2-locus population genetic model of insecticide resistance (Levick et al. in press) to illustrate the mechanisms by which these parameters are expected to influence the development of insectide resistance under alternative insecticide use strategies. Levick et al. (in press) describe the technical details of the model and explore large parameter ranges. We provide an accessible summary of the model and use selected parameter values to develop a mechanistic understanding of how inputs influence the development of resistance. This mechanistic understanding can provide a more robust framework for the debate on the relative merits of different insecticide use strategies.

@@

@@

@@

The simulation proceeds through generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness. This acts to make the fitter alleles more common over time. Thus a summary of the selection stage is :

> selection coefficient = resistance restoration * effectiveness  
>
> Locus fitness : 	
> SS = 1-effectiveness  
> RS = 1-effectiveness + (dominance*selection coeff.)  
> RR = 1-effectiveness + selection coeff.  
> 
> Allele fitness :   
> SS1SS2 = SS * SS  
> SS1RS2 = SS * RS  
> … and for other combinations  
> RR1RR2 = RR * RR  
> 
> Individual fitness = Allele fitness * exposure to insecticide  
> 
> In each generation :   
> Allele frequency[t+1] = Allele frequency[t] * Individual fitness

Sexual reproduction is represented first by creating gametes by splitting the adult genomes in two, and secound by forming offspring genotypes by adding gametes with recombination.

@@

Here we principally look at model outputs of the change in allele frequencies over time (generations). We investigate three insecticide use strategies.

1. single insecticide
1. two insecticides used in sequence, replacing the first with a second once the proportion of resistance alleles reaches 50%
1. two insecticides used in a mixture (concurrently)

@@

@@

#### results: Exposure
Increasing exposure to 0.8 from the base scenario reduces time-to-resistance for both mixture and sequential scenarios (Fig 8A relative to 6A). The change in exposure has a greater effect on the mixture relative to the sequential strategy and this results in a greater relative benefit for sequence versus mixture.

Similarly increasing exposure reduces time-to-resistance more for the mixture relative to sequence in the scenario in which one insecticide effectiveness has been increased (compare Fig 8B to 6B). The result is that the advantage of mixtures over sequence is removed by the increase in exposure.

@@

When single insecticide use was represented (Figs 3-5) in all cases the initial increase in resistance was most rapid with straight lines on the log plots up to around 50% followed by a slowing rate of increase up to 100%. Higher values of the five inputs investigated all led to more rapid increases in resistance.

For single insecticide use resistance responded identically to changing exposure and effectiveness (compare Fig 3A to 3B). This makes sense as vector kill is effectively a product of exposure times effectiveness. For example exposing 50% of a  population to an insecticide which is 75% effective would be expecetd to have the same result as exposing 75% of the population to one which is 50% effective. This observation points to the mechanism by which increasing both exposure and effectiveness lead to a faster increase in resistance. In both cases the increased deaths of susceptible vectors cause a higher selection pressure that can explain the faster increase.

@@

In comparison with the use of a single insecticide, using two insecticides either in sequence or in mixtures differs in some ways and where it differs that can again help us to understand the mechanisms. Looking at our base scenario (Fig 6A) the the curves for resistance to both insecticides in the mixture are shallower than those for sequential use and in this particular example are identical because they have set the same parameter values for each insecticide to be the same. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Thus each insecticide reduces the selection pressure for the increase in resistance to the other and could be said to 'protect' the other. However this protection is not necessarily enough to ensure that resistance arises more slowly in a mixture relative to a sequential strategy. As shown in our base scenario, resistance can arise more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other.

@@

**Ian, Matt this is still the bit I can't quite explain. Any suggestions ?**

The biggest difference I can't explain is why that in mixtures, increasing the exposure to both insecticides results in a decrease in time-to-resistance (from 90 in Fig 6A to 40 in Fig 8A) and yet increasing the effectiveness of both insecticides by the same amount results in a slight increase in time-to-resistance (from 90 in Fig 6A to 100 in Fig 7). This contrasts with the identical effects that exposure and effectiveness have on a single insecticide (Figs 3 A,B) and on insecticides in sequence (compare Figs 7 & 8A).

either

1. Why doesn't increased exposure provide 'protection' to the other insecticides in a mixture ?
1. Why doesn't increased effectiveness lead to a greater selection for resistance in each insecticide ?
1. Why do exposure & effectivness have the same effect on a sequence but different effects on a mixture ?

*some previous guesswork I'm now not sure about: Increasing exposure gives shorter time-to-resistance for both mixtures and sequences due to increased deaths of susceptibles creating a higher selection pressure. In a mixture the effect of exposure is greater and this means that increasing exposure can make sequences preferable to mixtures. The reason for the greater effect of exposure on mixtures could be that exposure is having both a direct effect on each insecticide and an indirect effect on the other. Thus higher exposures lead to faster resistance in each insecticide (direct), in turn increased resistance to each insecticide reduces the 'protection' that it gives the other, leading to faster resistance in the other (indirect). This could act as a postive feedback loop where faster resistance in one contributes to faster resistance in the other.*

Mechanisms for the response of resistance in mixtures are summarised in Table 3. Table 4 summarises mechanisms explaining the difference between mixture and sequential strategies.

@@

@@

(@) exposure ...

(@) ? In a mixture the effectiveness of one insecticide has major effects on the resistance of the other. In contrast the dominance and selective advantage of one insecticide has less effect on the resistance of the other. For exposure we kept values the same for each insecticide (to represent a mixture in which the 2 insecticides are applied together). However we predict that if they were applied separately then (similar to effectiveness) the level of exposure to one insecticide would influence the time-to-resistance of the other.

@@

1. In discussion and maybe elewhere address potential concern that the relationships shown here may not hold for all of paramter space.
1. can I develop explanatory diagrams to describe the input parameters ?
1. mention generation time of 1 month
1. mention other potential effects e.g. differential exposure of M&F, sex linkage, ...
1. add to discussion about relevance of these mechanisms in the field
1. e.g. exposure can be reduced if either insecticide has a repellant effect, which can be set in formulation. (Birget & Koella 2015 mention this not for resistance but for ITNs with repellancy requiring higher coverage for elimination)
